---
title: "Introduction to R"
author: "Andrew McAdam and Karl Cottenie"
date: '2018-09-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\tableofcontents  
```{r packages, echo=F, message=FALSE}

```

# Welcome
Welcome to R!  This document will provide you with an introduction to R.  If you follow along and perform the various tasks as you move through the document you will get a decent introduction to R.  Feel free to fiddle around as you go.  Making mistakes is a great way to learn!!  Content in this document was originally written as a script for lectures in Andrew's ZOO851 class at Michigan State University so some old terminology might persist.  Karl has also added a bunch of new content, but I (Andrew) take responsibility for all mistakes and appreciate Karl's expertise in make this document better and more comprehensive.  Please let me know if you see errors or if there are parts that aren't clear.  I will try and improve it as a "working document" for future classes.

In this document you will find R code as well as the output from R as it will appear on your screen. This
document will provide in a lot of cases both the input and output of the R session. I recommend that as you work your way through the document, you type the code into R, instead of using copy-paste. This will allow you to get used to typing R code, and get used to trouble-shooting any error messages you will get.


I will provide both a pdf copy of this document, but it has been created using a free software program called Rstudio (http://rstudio.org).  This program allows you to create PDFs like this of your work in R using a .tex format (i.e. LaTex).  RStudio is a very useful interface for R and producing PDF documents like this will allow you to keep very careful notes about what you have done and be able to communicate these (to your supervisor, perhaps) so that it is perfectly clear what you have done.  As a supervisor, I require all of my students to leave me with an archive folder of all of their data and R code.  Having a document like this would then allow me to see exactly how they performed their analyses and to recreate their analyses.  This sort of paper trail is essential for the reproducability of scientific findings and will save a lot of time in the long-run if you get used to working in this format. 

# Interactive Interface
## Interacting with R


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
